---
title: "Home Ownership Rates in the United States Between Generations"
author: "Kyle Vavasour"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

## Introduction

In recent years there has been an increasing pessimism about the prospect of 
owning a home, with younger generations dissatisfied how older generations
own homes while they cannot afford to. Younger generations may question whether
they will ever be able to own their own homes, and compare themselves to 
the seemingly far favorable conditions for older generations; indeed, home 
ownership and housing affordability is often cited as a core issue for
younger generations [3].

\
The goal of this project is to 
examine differences in home ownership between generations through home ownership rates by 
year, age, and locations. The proposed research question is: Have home ownership rates in 
the United States changed between the generations of Baby Boomers, Gen X, Millennials, 
and Gen Z?

\
Answering this question is not as simple as comparing home ownership rates directly
between generations; as we might expect, generations that have been alive longer will
tend to have higher rates due to more time to gain wealth and more time in the housing
market. Instead, to better compare home ownership rates, we have to look at the trends
between generations of time, age, and geography to see if there are real differences.
Younger generations cite home ownership as a core issue, but to get a sense of
whether home ownership has truly declined, an analysis of the trends over age, 
time, and geography must be preformed. Conducting this analysis will provide
useful insight into the issue of home ownership between generations.

\
The dataset that was used to answer this question is a subset of the 
IPSUM CPS dataset. The dataset comes from the United States
Census Current Population Survey (CPS) and its supplemental Annual Social and 
Economic Supplement survey (ASEC). The CPS is preformed monthly to different
households in the US while the ASEC supplement is preformed annualy in March.
The dataset contains both household and individual level information, meaning
some variables refer to entire household information such as whether a house
is owned or rented, while other variables refer to individual information
such as a persons age.

\
As the entire IPUMS CPS dataset is large and contains many variables
that are irrelevant to this analysis, a subset of the data was used containing
only the variables and year range of interest described in the methods section.


## Methods

The methods section includes subsections corresponding to how the data was
obtained, data cleaning, data wrangling, methods for producing plots, and methods
for predictive modelling.

### Data Collection

IPUMS CPS allows users to create
their own subsets, called data extracts, of the full CPS dataset that contain
variables and year ranges of interest. To create a data extract, a user must
select variables and year ranges and then wait for a data extract to
be generated; once the extract is generated it can be downloaded. The creation
and retrieval of the data extract used in this analysis was preformed through
the IPUMS API. To create the extract through the API, a post request was made
using the httr R package with a formatted JSON string that specified the year range
and variables of interest. The year range was 1976-2023 and the variables of 
interest were:

\
-YEAR which is the year an observation was from

-AGE which is the age of an individual

-OWNERSHIP which is whether the household that an individual belongs to is rented/owned

-RELATE which is the relation to their relation to the householder of the household an individual belongs to

-STATEFIP which is the FIP code of the state were the observation was from

-COUNTY which is the FIP code of the county were the observation was from

-HHINCOME which is the total household income of the household an individual belongs to

\
Once the data extract request was submitted, a get 
request was used to check whether the data extract had been generated. Finally,
when the data extract was complete, the data files were downloaded from the
URLs provided from the content of the get request. The dataset was then loaded
in R using the ipumsr package provided by IPUMS for loading IPUMS data into
R.

\
In addition to the key variables mentioned in the introduction, IPUMS provides
weights for calculating individual and household level statistics to mitigate
the sampling bias of the the survey, these weights were also included in the
dataset as the variables ASECWTH for household level statistics
and ASECWT for individual level statistics.

Finally, more data was acquired for use in the predictive modelling section.
Historical average mortgage interest rates in the US between 1976-2023 were 
obtained from Federal Reserve Economic Data webiste [4]. Additionally, data 
on historical median home prices in the US between 1976-2023 was obtained from 
the DQYDJ website [5]. Finally, data for adjusting
IPUMS economic variables such as HHINCOME for inflation was obtained
from the IPUMS website which provided conversion rates for each year to standardize
the variables to the year 1999.


### Data Cleaning

Once the dataset was loaded into R, it's dimensions were checked for how
many observations were present in the dataset and whether the number of variables
was consistent with what was requested in the data extract. Checking the dimension
showed that the dataset contains 8,293,750 observations with each observation being an 
individual who belongs to a household. The head, tail, and variable types were also 
checked to make sure the dataset was imported correctly.

\
Once the dataset was confirmed to have been imported correctly, the variables
were checked for missing values and suspicious values. All values for AGE,
YEAR, RELATE, OWNERSHP, STATEFIP, HHINCOME, and INCRETIR had no missing values and were all in
their expected range; however, the variable COUNTY had 3,159,092 missing values
and 2,991,865 values of 0 which is not a recognized county FIP code. Checking
the IPUMS data dictionary confirmed that a value of 0 was used for counties that
did not meet a population threshold due to concerns about deanonymizing
data. Further investigation also showed that all the missing values in the COUNTY
variable occurred before 1995. Removing observations with missing values for
COUNTY would remove a significant amount of the data. Additionally, the COUNTY
variable was not used across most of the analysis; therefore, no observations
were removed.

\
Something to note in the dataset is that the values in the categorical variables
are given by codes that correspond to their meaning, for example, the variable
OWNERSHIP has a value of "10" when the observation belongs to a household that 
is owned. The meaning of these codes is easy to find on the IPUMS CPS website,
so these values were not changed in this analysis. The additional data aqquired
on median home prices was already adjusted for inflation, so no further adjustments
were required.

### Data Wrangling

The number of observations with age >90 were small which created outliers in
the home ownership rates within these age ranges. Additionally, In the United States, 
people cannot own property until they reach the age of majority which is 19 in some states. 
Therefore, the age ranges of <19 and >90 were removed from the data set. After removing
all observations with age less than 19 and greater than 90 there were 5,796,909 observations left.
Removing these entries means that the population of interest in this analysis are
adults in the United States between the ages of 19 and 90.

\
For the analysis, a new variable had to be created referring the generation that individual belongs to. 
There are no official age cutoffs for what defines a generation, so instead
the generation age ranges were chosen to be consistent with an 
article by Pew Research Center, the age ranges were: 1946-1964 (Baby Boomer),
1965-1980 (Gen X), 1981-1996 (Millennial), 1997-2012 (Gen Z) [2]. The generation
variable was created by subtracting the AGE variable from the YEAR variable to
get a birth year and then checking which generation range it was in. Finally,
the frequencies of each generation was checked to see if the variable was
create properly, as you might expect, Gen Z had the least amount of observations 
at 86,029 while Baby Boomers had the most 2,147,536	due to the data set containing
many years before any individual in Gen Z was born. 

\
For the predictive modelling, HHINCOME had to be adjusted for inflation
so comparisons between observations would be accurate. To do this, the inflation
conversion data acquired from IPUMS was used; the data provided a conversion
rate for each year from 1976-2023 that could be used to standardize the income
data to dollars in the year 1999. Using these conversion rates, the variables
HHINCOME and INCRETIR were multiplied by the rate corresponding to the year that
the observation was collected in.


### Methods for Plots

Once the data cleaning and wrangling was complete, the analysis could be preformed. 
To answer the research question, home ownership rates had to be calculated across generations. 
For this analysis, a home ownership rate was defined:

$$\text{Home Ownership Rate (\%)} = 100 \cdot \frac{\text{Owner Occupied Households}}{\text{Total Occupied Households}}$$

\
This definition is consistent with the definition given by the Census Housing Vacancy
Survey [2]. As the dataset provides weights for calculating household level
statistics, the home ownership rate was calculated as the the sum of the weights
associated with an owner occupied household divided by the sum of weights associated
with an occupied household. Finally, a generations home ownership rate is
calculated the same, except only households owned/occupied by an individual in
that generation are counted.

\
To find owner occupied households within the dataset both the OWNERSHP and RELATE variable had
to be used. The OWNERSHP variable in the dataset refers to whether the household the observation
belongs to is owned or rented; this does not mean that the individual is a 
homeowners as, for example, an individual living with parents who own a home would
have a value of ownership rather than rented in the OWNERSHIP variable. 
Instead, both the variables OWNERSHP and RELATE have to be used to find homeowners. 
If OWNERSHP indicates that the observation
belong to a household that owns their home and RELATE indicates that the observation
is the householder then that observation is considered a homeowner.

\
Finally, visualizations and summaries were created to explore the differences in 
homeownership rates between generations. 

\
First, table summaries were created
to show homeownership rates by generation in different years and at different ages. 
To create the table of homeownership rate by year between generations, homeownership
rates were calculated for each year grouped by generation. To create the table of
homeownership rate by age between generations, homeownership rates
were calculated for each age grouped by generation; for example, the homeownership
rate for 25 year old Baby Boomers would be calculated using only observations
were the age was 25 and the observation was a Baby Boomer (these would be from
years between 1971-1989).

\
To further explore the trend of homeownership rate over time, a line chart was created to show 
homeownership by year with a trend line for each generation. 
Similarly, to explore the trend of homeownership by age, a line chart was created 
to show homeownership by age with a trend line for each generation.

\
To explore difference in homeownership in locations by generation, a map visualization
was created to show home ownership at the state level, and at the county level. To
create these map visualizations, geojson files of USA state and county maps 
that included FIPS data were used. Home ownership rates were calculated by
state/county and then this data was merged into the geojson files using the
FIPS codes. For the county map, some counties were not identified, these
counties were merged together within state borders to form a map that 
contained homeownerhip rates in all identified counties, and home ownership
rates in states not including identified counties within the state. Similarly,
any county with <50 observations in any generation was combined with the
non-identified counties within state borders.

### Methods for Modelling

The goal of this section is to model home ownership rates over time between
generations, ideally creating a model capable of predicting home ownership trends
into the near future. This is a complicated task as home ownership rates likely depend
on a large amount of external factors such as housing market conditions, economic
conditions, and social factors. While the IPUMS CPS data set provides some measures
that may be useful in prediction, such as household income and retirement income, the
variables available are limited. To compensate for some missing variables, data for
mortgage interest rates and median home prices by year was merged into the data set.

\
There are also a variety of ways to approach modelling home ownership by
generation. One possibility is to model home ownership at the household level; this would
involve creating a model that predicts whether a specific household is owned or
not (or probability that a specific household is owned) 
using household level data such as the homeowners income and home prices
within the area. An advantage to this approach would be the ability to incorporate
specific data such as geography and age of homeowner. Another approach is to model home ownership rates
directly, predicting population level home ownership for a given year 
between generations based on population level information such as 30 year
fixed interest rates for mortgages and mean income for people within the generation. An advantage
to this approach is that it directly models the outcome, that being home ownership rates for a given year and generation, making it more effective at predicting future home ownership rates. For this analysis,
the second approach was chosen due to the model using relatively simpler population level
data compared to the more complex household level information making it easier to apply in predicting
future trends.

\
An initial GAM was fit using home ownership rate as the outcome with year as a 
smooth term and generation as a factor for smooth interactions; this results
in separate smooths fit for each generation that share a smoothing parameter. For the
smooth term, 8 basis functions were used and a second derivative penalty was also used.
A plot for the GAM was produced and compared to the observed trend to assess the
fit of the model. Next, another GAM was fit using the same parameters but withholding
the last 5 years of data within the data set to test how well the model extrapolates to
trends five years in the future. 

\
Next, a second GAM was fit using home ownership rate as the outcome with 
average household income and average retirement income as smooth terms with generation
as a factor for smooth interactions as well as 30-year fixed mortgage rates and median
home price as smooth terms without interaction. For the income smooth term, 3 basis
functions were used and a first and second derivative penalty were used. For the
home price and mortgage rate smooth terms, 10 basis functions were used with a 
second derivative penalty. A plot for the GAM was produced and 
compared to the observed trend to assess the fit of the model. Next, another GAM was
fit using the same parameters but withholding the last 5 years of data within the
data set to test how well the model extrapolated trends five years in the future.

\
Summaries of the models where produced to assess the significance of the terms
and obtain deviance explained and r-squared values. For the models with withheld
data, mean absolute error was calculated within the year ranges that were withheld.
The five year projections of both GAMs trained with withheld years were compared
against the true five year trend to assess reliability and accuracy.


```{r warning=FALSE, echo=FALSE, output='none', message=FALSE}
library(ipumsr)
library(raster)
library(tidyverse)
library(gridExtra)
library(httr)
library(leaflet)
library(kableExtra)
library(mgcv)
```

```{r eval=FALSE, echo=FALSE}
IPUMS_API_KEY = "[API KEY]"
```

```{r echo=FALSE, eval=FALSE}
# create body of post request as JSON string
body <- '{
    "description": "API Extract",
    "data_structure": { 
        "rectangular": {
            "on": "P"
        }
    },
    "data_format": "fixed_width",
    "variables": {
      "YEAR":{},
      "AGE":{},
      "ASECWTH":{},
      "RELATE":{},
      "OWNERSHP":{},
      "STATEFIP": {},
      "COUNTY": {},
      "HHINCOME": {},
      "INCRETIR": {}
    },
    "samples": {'
# loop over years to add them all to "samples" part of body
for (year in 1976:2022) body <- paste0(body, "\"cps", year, "_03s\"", ":{},")
body <- paste0(body, "\"cps2023_03s\"", ":{}}}")

# post request with body
post_request <- POST(
  url = "https://api.ipums.org/",
  path = "extracts",
  query = list(
    "collection" = "cps",
    "version" = "beta"
  ),
  config = add_headers(
    "Authorization" = IPUMS_API_KEY,
    "Content-Type" = "application/json"
  ),
  body = body
)
```

```{r eval=FALSE, echo=FALSE}
# repeat checking if the request is complete
get_content <- NULL
time_between_check <- 60
repeat{
  # get request
  get_request <- GET(
    url = "https://api.ipums.org/",
    path = "extracts",
    config = add_headers(
      "Authorization" = IPUMS_API_KEY,
      "Content-Type" = "application/json"
    ),
    query = list(
      "collection" = "cps",
      "version" = "beta"
    )
  )
  
  # check if request is completed
  get_content <- content(get_request)[[1]] #get the most recent request content
  if (get_content$status == "completed") break
  # if not wait 30s and then try again
  Sys.sleep(time_between_check)
}

# get urls of xml and data
data_link <- get_content$download_links$data$url
xml_link <- get_content$download_links$ddi_codebook$url
 
# write .dat to directory
data_get <- GET(
  url = data_link,
  config = add_headers(Authorization=IPUMS_API_KEY),
  write_disk("data_file.dat.gz", overwrite=TRUE)
)

# write .xml to directory
xml_get <- GET(
  url = xml_link,
  config = add_headers(Authorization=IPUMS_API_KEY),
  write_disk("xml_file.xml", overwrite=TRUE)
)
```

```{r echo=FALSE, output='none', warning=FALSE, message=FALSE}
# Use IPUMS functions to get data table
ddi <- read_ipums_ddi("./data/xml_file.xml")
data <- read_ipums_micro(ddi, data_file="./data/data_file.dat.gz")
```

```{r eval=FALSE, echo=FALSE}
# check data dimension, head, tail, and variables
head(data)
tail(data)
dim(data)
str(data)
```

```{r eval=FALSE, echo=FALSE}
# key variables are AGE, RELATE, OWNERSHP, ASECWTH, YEAR
data <- data %>% select(AGE, RELATE, OWNERSHP, ASECWTH, YEAR, STATEFIP, COUNTY, HHINCOME, INCRETIR)
```

```{r eval=FALSE, echo=FALSE}
# key variables
summary(data$AGE)
summary(data$YEAR)
summary(data$HHINCOME)
table(data$RELATE)
table(data$OWNERSHP)
table(data$STATEFIP)
table(data$COUNTY)
```

```{r eval=FALSE, echo=FALSE}
# further investigation of county NA (when do they occur?)
data %>%
  filter(is.na(COUNTY)) %>%
  group_by(YEAR) %>%
  summarise(count=n()) %>%
  ggplot(aes(x=YEAR, y=count)) +
  geom_bar(stat='identity')
```

```{r eval=FALSE, echo=FALSE}
# check distribution of age
data %>%
  ggplot(aes(x=AGE)) +
  geom_histogram()
```

```{r eval=FALSE, echo=FALSE}
# check old outliers
data %>%
  filter(AGE > 90)
```

```{r, echo=FALSE}
# filter younger than 18
data <- data %>% filter(AGE >= 19 & AGE <= 90)
```

```{r eval=FALSE, echo=FALSE}
# check number observations after removing <18
dim(data)
```

```{r eval=FALSE, echo=FALSE}
# what is the earliest generation in the dataset?

min(data$YEAR - data$AGE) #earliest birthday

data %>% # table of counts by birthyear
  mutate(BIRTHYEAR = YEAR-AGE) %>%
  group_by(BIRTHYEAR) %>%
  summarise(n())

data %>% # counts for gen before silent and all after
  mutate(BIRTHYEAR = YEAR-AGE) %>%
  filter(BIRTHYEAR < 1946) %>%
  summarise(n())
```

```{r echo=FALSE}
# create the generation variable
data <- data %>%
  mutate(GEN = ifelse( (YEAR-AGE >= 1946) & (YEAR-AGE <= 1964), "Baby Boomer",
               ifelse( (YEAR-AGE >= 1965) & (YEAR-AGE <= 1980), "Gen X",
               ifelse( (YEAR-AGE >= 1981) & (YEAR-AGE <= 1996), "Millenial",
               ifelse( (YEAR-AGE >= 1997) & (YEAR-AGE <= 2012), "Gen Z", 
               "Pre 1946")))))
data$GEN <- factor(data$GEN, levels=c("Baby Boomer", "Gen X", "Millenial", "Gen Z", "Pre 1946"))
```

```{r eval=FALSE, echo=FALSE}
# check counts by generation
data %>% 
  group_by(GEN) %>%
  summarise(n())
```

## Analysis

The analysis section includes subsections corresponding to the initial analysis subsection
where plots where produced and analyzed and the modelling subsection where a predictive
model was created and assessed.

### Initial Analysis

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(YEAR %in% 2018:2023) %>%
  filter(RELATE == 0101) %>%
  mutate(Year = YEAR) %>%
  group_by(Year, GEN) %>%
  reframe(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  pivot_wider(names_from=GEN, values_from=RATE) %>%
  kable(caption="Home Ownership Rates (%) between 2018-2023 by Generation")
```

\
This table shows the homeownership rates by generation in recent years. Since
the older generations like Baby Boomers and Millennials have been alive longer,
and likely have more wealth, we would expect to see that their home ownership
rate is much higher as shown in the table. We also see that the home ownership
rate for Millennials has increased significantly in this time period, around 
a 14% difference with a signicant increase in 2020. Most Millennials would be
in their late 20s and early 30s which is around the age of the average first
time homebuyer in the United States. There were also historically low interest rates
between 2019-2021 which could have contributed to this increase. Finally, we
can also see that Baby Boomers were the only generation to have a decrease in
homeownership rates within this time period, this could be due to aging populations
selling their home to move in with family or into care homes. In the future, we
would expect to see the Baby Boomer home ownership rate continue to fall. Overall,
we see that Millennials and Gen Z have the largest increase in home ownership
rates within this time period.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(AGE %in% seq(from=25, to=40, by=5)) %>%
  filter(RELATE == 0101) %>%
  mutate(Age = AGE) %>%
  group_by(Age, GEN) %>%
  reframe(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  pivot_wider(names_from=GEN, values_from=RATE) %>%
  kable(caption="Home Ownership Rates (%) for ages 25-40 by Generation")
```

\
This table shows the homeownership rates at certain ages by generation. As the
oldest members of Gen Z were only 26 in 2023, their is no data available in
the later ages. We can see that Baby Boomers have the highest home ownership
rate at all ages while Millennials have the lowest across all ages except
25. While Gen Z only has data at the age 25, it has the second highest home
ownership rate at this age. There does appear to be a trend of homeownership
rates dropping at all ages as the generations get younger; however, Gen Z could
disrupt this trend if their rates continue to be high. Without further data on
Gen Z, we cannot know if this trend will continue or Gen Z will outperform 
Millennials and/or Gen X.

```{r echo=FALSE}
# over all age groups
rate_by_time_all <- data %>%
  group_by(YEAR) %>%
  filter(RELATE == 0101) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  drop_na() %>%
  ggplot(aes(x=YEAR, y=RATE)) +
  geom_line(linewidth=1, color='black') +
  ylim(0, 100) +
  labs(y="Home Ownership Rate (%)", x="Year", title="Figure 1. 
Home Ownership Rate By Year")

rate_by_age_all <- data %>%
  group_by(AGE) %>%
  filter(RELATE == 0101) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH)) %>%
  drop_na() %>%
  ggplot(aes(x=AGE, y=RATE)) +
  geom_line(linewidth=1, color='black') +
  ylim(0, 100) +
  labs(x="Age", title="Figure 2. 
Home Ownership Rate By Age") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

grid.arrange(rate_by_time_all, rate_by_age_all, ncol=2)
```
\
These plots show overall trends of home ownership by year and age. We can see
that home ownership rates tend to stay fairly similar over time. This makes sense
as outside of large economic shifts we would expect to have the proportion of
people owning homes stay fairly constant as younger people buy houses and older
people move out.We can also see that home ownership rates by age increases over 
time and then starts to flatten out around 50, before starting to decrease at 
around 80. Again, this would make sense as people would want to buy homes once
they have a stable income and career which would likely occur in the age range
25-40. Then since most people who had the means to buy a house or wanted to buy
a house likely already did by 50, the rate starts to flatten. Finally, once
a person is older, they may start to sell their homes and move in with family
or care home. Interestingly, we also see an outlier at around 19 where the rate 
starts off higher. The slightly elevated rate at 19 could be due to inheritance where 
an individual is not legally allowed to own a property until they reach age of majority.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE == 0101) %>%
  group_by(YEAR, GEN) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH), .groups="rowwise") %>%
  drop_na() %>%
  ggplot(aes(x=YEAR, y=RATE, color=GEN)) +
  geom_line(linewidth=1) +
  ylim(0, 100) +
  theme(legend.position="bottom") +
  labs(x="Year", y="Home Owneship Rate (%)", title="Figure 3. Home Ownership Rate by Year and Generation",
       color="Generation:")
```
\
This plot shows the trend of home ownership by year between each generation.
We can see that Baby Boomer rate is has flattened significantly since
around 2005 and the Gen X was starting to flatten around 2005 as well
although it has started to increase again since around 2015; the sub prime mortgage
crisis occurred between 2007 and 2010 and could have contributed to this decrease
in the rate of change of the Gen X home ownership rate. For Millennials, we can 
see that the rate has been increasing and is starts to increase faster around
2015. Finally, for Gen Z there is not a lot of data, but we can see that it
is starting to increase and in 2020 it had the rate started to increase faster.

```{r echo=FALSE}
data %>%
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE==0101) %>%
  group_by(AGE, GEN) %>%
  summarise(RATE = 100*weighted.mean(OWNERSHP==10, ASECWTH), .groups="rowwise") %>%
  drop_na() %>%
  ggplot(aes(x=AGE, y=RATE, color=GEN)) +
  geom_line(linewidth=1) +
  ylim(0, 100) +
  theme(legend.position="bottom") +
  labs(x="Age (Years)", y="Home Owneship Rate (%)", title="Figure 4. Home Ownership Rate by Age and Generation",
       color="Generation:")
```

\
This plot shows the trend of home ownership by age between each generation.
The plot provides a good comparison of home ownership rates by generation.
We can see that Baby Boomers have consistently outperformed all other generations
past the age of 26 and we can also see that Millenials have under preformed all
other generations past the age of 26. The Gen X rate tracks fairly closely to the
Baby Boomer rate before it starts to under preform in the 30s age range and onward.
Gen Z out preforms both Millennial and Gen X for the age range where data is available,
while the Gen Z line flattens towards the age of 26, without further data we do not
know whether Gen Z will outperform or under preform other generations. Low intrest
rates from 2019-2021, when older members of Gen Z would have started thinking about
buying a home, could have contributed to the high home ownership rates of 
Gen Z relative to the other generations.

\
To see the interactive visualization of home ownership rates by state 
between generations that is analyzed in the following paragraph click [here](https://kylevav.github.io/JSC370-Final-Project/visuals.html#Home_ownership_rates_by_state_and_generation).

\
The map shows home ownership rates by state and generation. We can see that 
Baby Boomers have fairly high home ownership rates across all states. The 
states with the lowest home ownership rates for Baby Boomers are California,
New York, and Hawaii. For Gen X, we see lower home ownership than Baby Boomers.
We also see that Midwestern states and non-coastal states tend to have higher
home ownership rates, which could be due to affordability. For Millennials, we 
can see again that home ownership is lower than Gen X, but we also see that 
the states with the lowest home ownership rates are similar to the states with 
the lowest home ownership rates for Gen Z. Finally, for Gen Z the states with
the lowest home ownership rates are fairly similar to Millenials and Gen X, 
but surprisingly Florida is one of the states with the highest home ownership
rate which is unique to Gen Z.

\
\
To see the interactive visualization of home ownership rates by county 
between generations that is analyzed in the following paragraph click [here](https://kylevav.github.io/JSC370-Final-Project/visuals.html#Home_ownership_rates_by_county_and_generation).

\
Similar to the previous map, this map shows home ownership rates by
geography and generation. This map includes counties that were identified
in the dataset, many counties were not identified in the dataset as their
population was below a threshold were there are concerns about deanonymizing
the data. All counties that were not identified in the dataset and any
county with less than 50 observations in any generation are combined 
into the state and labeled "State w/o Counties". Similar patterns can be seen
to the previous state map; however, the difference between urban and rural
home ownership rates are more obvious in this map, with rural areas tending to
have higher home ownership rates across all generations.

### Modelling

As described in the methods section, the goal of this section is to create a model
that can predict the home ownership rates of a generation at a given year; the trend
that we want to model can be seen in figure 3. The first GAM used year with an
interaction for generation as predictors.

```{r echo=FALSE}
# additional variables

# https://fred.stlouisfed.org/series/MORTGAGE30US
intrest_rates <- read_csv('./data/MORTGAGE30US.csv')
intrest_rates <- intrest_rates[grepl('-03-', intrest_rates$DATE),]
intrest_rates$DATE <- as.numeric(format(as.Date(intrest_rates$DATE, format="%d-%m-%Y"),"%Y"))
intrest_rates <- intrest_rates %>%
  group_by(DATE) %>%
  summarise(INTREST = mean(MORTGAGE30US))

inflation <- read_csv('./data/inflation.csv')
inflation$year <- as.numeric(inflation$`Data Year`)-1
inflation$multiply <- as.numeric(inflation$CPI99)
inflation <- inflation %>%
  select(year, multiply) %>%
  drop_na()

#https://dqydj.com/historical-home-prices/
housing_prices <- read_csv('./data/homeprice.csv')
housing_prices <- housing_prices[grepl('Mar', housing_prices$category),]
housing_prices$year <- 1953:2023
housing_prices <- housing_prices %>%
  select(year, `CPI-Adjusted Price`) %>%
  filter(year >= 1976)
```

```{r echo=FALSE}
annual_rates <- data %>%
  left_join(inflation, join_by(YEAR == year)) %>%
  left_join(intrest_rates, join_by(YEAR == DATE)) %>%
  left_join(housing_prices, join_by(YEAR == year)) %>%
  mutate(HHINCOME = HHINCOME * multiply,
         INCRETIR = INCRETIR * multiply) %>%
  filter(GEN != 'Pre 1946') %>%
  filter(RELATE == 0101) %>%
  group_by(YEAR, GEN) %>%
  summarise(RATE = weighted.mean(OWNERSHP==10, ASECWTH),
            INCOME = weighted.mean(HHINCOME, ASECWTH),
            RETIREMENT = weighted.mean(INCRETIR, ASECWTH),
            INTREST = weighted.mean(INTREST, ASECWTH),
            PRICE = weighted.mean(`CPI-Adjusted Price`, ASECWTH),
            WEIGHT = n(),
            .groups="rowwise") %>%
  drop_na()

annual_rates_train <- annual_rates %>%
  filter(!(YEAR %in% c(2023, 2022, 2021, 2020, 2019)))

annual_rates_withheld <- annual_rates %>%
  filter(YEAR %in% c(2023, 2022, 2021, 2020, 2019))
```

```{r echo=FALSE}
knots <- list(x = c(1976, 2023, 2030))
gam_year <- gam(RATE~GEN+s(YEAR, GEN, bs='fs', k=8, m=2), data=annual_rates, method = "REML",
         knots=knots, family=binomial(link="logit"), weights=WEIGHT)
gam_year_withheld <- gam(RATE~GEN+s(YEAR, GEN, bs='fs', k=8, m=2), data=annual_rates_train, method = "REML",
         knots=knots, family=binomial(link="logit"), weights=WEIGHT)
```

```{r echo=FALSE, eval=FALSE}
summary(gam_year)
summary(gam_year_withheld)
```

```{r echo=FALSE}
gam_summary <- summary(gam_year)
gam_summary$p.table %>% kable(caption="GAM fixed terms")
gam_summary$s.table %>% kable(caption="GAM smooth terms")
```

```{r echo=FALSE}
crit <- qnorm((1 - 0.95) / 2, lower.tail = FALSE)
pred <- as.data.frame(predict(gam_year, annual_rates, se.fit = TRUE, type="response"))
pred$upr_ci <- pred$fit + (crit * pred$se.fit)
pred$lwr_ci <- pred$fit - (crit * pred$se.fit)
plot_data <- cbind(annual_rates, pred)

plot_data %>%
  ggplot(aes(x=YEAR)) +
  geom_ribbon(aes(ymin = 100*lwr_ci, ymax=100*upr_ci, fill=GEN), alpha=0.2) +
  geom_point(aes(y=100*RATE, color=GEN), size=1.5) +
  geom_line(aes(y=100*fit, color=GEN), linewidth=1, alpha=1) +
  ylim(0, 100) +
  theme(legend.position="bottom", legend.title=element_blank()) +
  labs(x="Year", y="Home Owneship Rate (%)", title="Figure 5. Modelled Home Ownership Rate by Year and Generation")
```

\
This plot shows the fit of the GAM using year as a smooth term with generation
as a smooth interaction. The observed ownership rates are shown as the points
on the graph. The model achieves a very good fit to the data with 99.7% deviance
explained and 0.996 adjusted R-squared. This is not unexpected as the model is
producing almost an exact fit to the observed data; using this model beyond the
years collected in the data set will likely produce unreliable predictions, especially
as the year gets further from the observed years in the dataset. However, the model 
does show that both the generation intercepts and the smooth term
where significant at p<0.001 showing that there is likely significant
differences between the trends of home ownership between generations.

\
While the GAM will likely be unreliable far outside the 1976-2023 year range,
we can test how well the model preforms on predicting home ownership trends
five years into the future by withholding the last five years (2019-2023) in the 
dataset and training the model only on the remaining years.

```{r echo=FALSE}
crit <- qnorm((1 - 0.95) / 2, lower.tail = FALSE)
pred <- as.data.frame(predict(gam_year_withheld, annual_rates, se.fit = TRUE, type="response"))
pred$upr_ci <- pred$fit + (crit * pred$se.fit)
pred$lwr_ci <- pred$fit - (crit * pred$se.fit)
plot_data <- cbind(annual_rates, pred)

plot_data %>%
  ggplot(aes(x=YEAR)) +
  geom_ribbon(aes(ymin = 100*lwr_ci, ymax=100*upr_ci, fill=GEN), alpha=0.2) +
  geom_point(aes(y=100*RATE, color=GEN), size=1.5) +
  geom_line(aes(y=100*fit, color=GEN), linewidth=1, alpha=1) +
  ylim(0, 100) +
  theme(legend.position="bottom", legend.title=element_blank()) +
  labs(x="Year", y="Home Owneship Rate (%)", title="Figure 6. 2019-2023 projection compared against true 5 year trend") +
  geom_vline(xintercept=2018, linetype="dotted")
```

```{r echo=FALSE}
annual_rates_withheld$pred <- predict(gam_year_withheld, annual_rates_withheld, type="response")
mae <- mean(abs(annual_rates_withheld$RATE-annual_rates_withheld$pred))
```

\
This plot shows the GAM that was trained without the last five years (2019-2023)
and then shows what the model predicts the trend to be in this year range.
The true trend is shown as points. Within the year range 2019-2023, the model has
a mean absolute error of 0.0143 (where the home ownership rate is on the scale 0-1).
We can see on the plot that for Baby Boomers and Gen X, the mode ls predicted trend
is extremely similar to the observed trend; however, for Gen Z and especially
Millennial, the models predicted trend is not nearly as accurate. This is especially
notable in the Millennial predicted trend where the observed home ownership 
rate for 2020-2023 are outside the 95% credible interval for predicted values. This
implies that the model is not very accurate for predicting the five year trend for
millennial home ownership, but the model does accurately predict the trend for
Baby Boomers and Gen X.

\
While using only past data on home ownership rates between generation to predict
future home ownership rates between generations would make a simple model that could
be used to predict easily, it is unlikely
any model that does not incorporate more information would be able to make 
accurate predictions farther into the future. Instead, a model that uses 
broad information that may relate to home ownership rates such as 30 year fixed mortgage
rate, annual median home price, and average salary within a generation may be
provide a model that is better at predicting home ownership further into the future.


```{r echo=FALSE}
gam_full <- gam(RATE ~ GEN + s(INCOME, GEN, bs='fs', k=3, m=c(3, 2, 1, 0)) + s(PRICE) + s(INTREST), data=annual_rates, family=binomial, weights=WEIGHT)

gam_full_withheld <- gam(RATE ~ s(INCOME, GEN, bs='fs', k=3, m=c(3, 2, 1, 0)) + s(PRICE, k=10, m=2) + s(INTREST, k=10, m=2), data=annual_rates_train, family=binomial, weights=WEIGHT)
```

```{r echo=FALSE, eval=FALSE}
summary(gam_full)
summary(gam_full_withheld)
```

```{r echo=FALSE}
gam_summary <- summary(gam_full)
gam_summary$p.table %>% kable(caption="GAM fixed terms")
gam_summary$s.table %>% kable(caption="GAM smooth terms")
```

```{r echo=FALSE}
crit <- qnorm((1 - 0.95) / 2, lower.tail = FALSE)
pred <- as.data.frame(predict(gam_full, annual_rates, se.fit = TRUE, type="response"))
pred$upr_ci <- pred$fit + (crit * pred$se.fit)
pred$lwr_ci <- pred$fit - (crit * pred$se.fit)
plot_data <- cbind(annual_rates, pred)

plot_data %>%
  ggplot(aes(x=YEAR)) +
  geom_ribbon(aes(ymin = 100*lwr_ci, ymax=100*upr_ci, fill=GEN), alpha=0.2) +
  geom_point(aes(y=100*RATE, color=GEN), size=1.5) +
  geom_line(aes(y=100*fit, color=GEN), linewidth=1, alpha=1) +
  ylim(0, 100) +
  theme(legend.position="bottom", legend.title=element_blank()) +
  labs(x="Year", y="Home Owneship Rate (%)", title="Figure 7. Modelled Home Owneship Rate by Year and Generation")
```
\
This plot shows the fit of the GAM using average income as a smooth term with generation
as a smooth interaction as well as median home price and 30-year fixed mortgage rate as 
smooth terms without interaction. To produce this plot, the average income for a generation, mortgage rate,
and median home price were used as predictors to calculate home ownership rate for each year in 1976-2023.
The observed ownership rates are shown as the points
on the graph. The model achieves a very good fit to the data with 99.5% deviance
explained and 0.993 adjusted R-squared. Unlike the previous models, this model
does not directly fit the trend over time, rather it uses information about interest rates,
home price, and income at a given time and predicts home ownership rate. Ideally, this
would produce a model that would be capable of predicting future home ownership rates if
data on future income, interest rates, and home prices was available.
Like the previous model, the generation intercepts and all smooth terms where
significant at p<0.001 showing that there is likely significant differences between
the trends of home ownership between generations.

\
Similar to the previous GAM, we can test how well the model preforms on predicting
home ownership trends five years into the future by withholding the last five
years (2019-2023) in the dataset and training the model only on the remaining years.


```{r echo=FALSE}
crit <- qnorm((1 - 0.95) / 2, lower.tail = FALSE)
pred <- as.data.frame(predict(gam_full_withheld, annual_rates, se.fit = TRUE, type="response"))
pred$upr_ci <- pred$fit + (crit * pred$se.fit)
pred$lwr_ci <- pred$fit - (crit * pred$se.fit)
plot_data <- cbind(annual_rates, pred)

plot_data %>%
  ggplot(aes(x=YEAR)) +
  geom_ribbon(aes(ymin = 100*lwr_ci, ymax=100*upr_ci, fill=GEN), alpha=0.2) +
  geom_point(aes(y=100*RATE, color=GEN), size=1.5) +
  geom_line(aes(y=100*fit, color=GEN), linewidth=1, alpha=1) +
  ylim(0, 100) +
  theme(legend.position="bottom", legend.title=element_blank()) +
  labs(x="Year", y="Home Owneship Rate (%)", title="Figure 8. 2019-2023 projection compared against true 5 year trend") +
  geom_vline(xintercept=2018, linetype="dotted")
```

```{r echo=FALSE}
annual_rates_withheld$pred <- predict(gam_full_withheld, annual_rates_withheld, type="response")
mae <- mean(abs(annual_rates_withheld$RATE-annual_rates_withheld$pred))
```

\
This plot shows the GAM that was trained without the last five years (2019-2023)
and then shows what the model predicts the trend to be in this year range.
The true trend is shown as points. Within the year range 2019-2023, the model has
a mean absolute error of 0.0081 (where the home ownership rate is on the scale 0-1).
We can see on the plot that for all generations, the models predicted trend is 
quite accurate to the observed data. The only observed home ownership rate
that falls outside the models 95% credible interval is the prediction for
2020 for Generation X; all other predictions are very close to the predictive
trend or within the 95% credible interval of the predicted trend. We can see that
this model is more reliable in predicting future trends than the simpler model,
that only uses year and generation, by comparing the fit shown in the plot and
the mean absolute error values. However, this model is much more difficult to use
in aqquiring prediction for the future as to do so you would have to already
have predictions for average income, mortgage rates, and median home prices; aqquiring
data for future values of these variables is a challenging problem on its own. While
this model may be more difficult to use in prediction, the accuracy of its future 
trend prediction shows that interest rate, household income, and home price, and
generation are all predictive of home ownership rates.

\
Comparing between the two models, the first model was slightly better at fitting the
true trend within the year ranges that were in the training data; however, for
year ranges outside of the training data, the second model that used broader
housing information gave more accurate predictions in the withheld five year
range. While the first model could be used to easily predict future home ownership
trends, the accuracy of these predictions, especially as the predicted years
go far beyond the years used in training, will become unreliable. The second
model could potentially be used to predict future home ownership trends, but
this may be difficult as information on future mortgage rates, income within
a generation, and home prices would be neccessary. Overall, both models show
that home ownership trends between generations are different, and that 
broad housing information can be used to gain fairly accurate predictions of
home ownership rates; particularly, average income within a generation, mortgage
rates, and median home price are all predictive of the home ownership rate for
a generation in a year.

## Conclusion

The analysis compared home ownership rates across generations by investigating
trends in home ownership rate over time, age, and geography. 

\
The trend of home ownership rate over years between generations was similar,
with fast growth followed by a flattening of the curve; however, some
differences between home ownership rate over year trend could be seen, such 
as the increase in Gen X home ownership rate slowing down between the years
2005 and 2015, potentially due to housing market conditions in this period.
While some differences were present, the overall trends appeared similar, 
showing that the trend home ownership rate over time has not significantly 
changed between generations.

\
The trend of home ownership rate over age between generations was also similar.
Although some deviations could be seen in the trends between generations.
Millennials tended to have the lowest home ownership rates at all ages when
compared to the other generations, while Baby Boomers tended to have the highest
rates at all ages. Gen X tracked closely with Baby Boomer between the ages 
19-35 but started to under preform at later ages. Surprisingly, Gen Z was
had the highest home ownership rates in ages 19-23 and the second highest
in ages 23-26 with Baby Boomers as the highest in that age range. The trends
between each generation being similar and the high home ownership rates of Gen Z
indicate that home ownership has not increased or declined significantly between
generations despite Baby Boomers and Gen X tending to have higher rates.

\
The analysis of home ownership rates by state and by county showed that
home ownership rates tended to be similar across geography between the 
generations. Home ownership rates in less populous states and counties, such as
Midwestern states or non-identified counties, tended to be higher across all generations; 
although there were some interesting differences, such as Florida being one of the
states with the highest home ownership rate for Gen Z compared to all other 
generations were Florida was not one of the states with high home ownership rate.

\
The modelling of home ownership rates showed that past home ownership rate
trends over time could be used to predict future home ownership rates within
a five year time period, although these predictions were not as reliable for
the Millennial and Gen Z generations. Additionally, the modelling showed that 
average income within a generation, 30-year fixed mortgage rates, and median
home price were all predictive of home ownership rates within a generation, and
using them could provide very accurate predictions of future home ownership trends
if data on these variable was available for the years that we want to predict. Finally,
the modelling showed that there was significant differences in the trends between
generations confirming the initial analysis of the plot that showed small
differences in the trends such as the Gen X rate slowing down between 2005 and
2015.

\
Overall, the analysis showed the the general trends of home ownership rates
by year, age, and geography have remained similar between generations, although
there are significant differences in specific trends over time that can be seen in
the initial analysis and modelling. While home ownership rates do not appear to
have declined significantly between generations, there is a pattern of home ownership
trends flattening as the generations get younger. This analysis focuses on
home ownership rates which do not necessarily reflect general housing market
conditions, so it cannot give an accurate picture of the housing market or housing
affordability. While some additional information such as mortgage rates and home
prices was incorporated into the modelling, further analysis that included more
economic data would provide more cohesive information about housing market between
generations.

## Citations:

\
IPUMS:
Sarah Flood, Miriam King, Renae Rodgers, Steven Ruggles, J. Robert Warren,
Daniel Backman, Annie Chen, Grace Cooper, Stephanie Richards, Megan
Schouweiler, and Michael Westberry. IPUMS CPS: Version 11.0 [dataset].
Minneapolis, MN: IPUMS, 2023. https://doi.org/10.18128/D030.V11.0

\
Generation Ranges:

[1] Dimock, M. (2019, January 17). Defining generations: Where millennials end and generation Z begins. Pew Research Center. https://www.pewresearch.org/short-reads/2019/01/17/where-millennials-end-and-generation-z-begins/  

\
Home Ownership Definition:

[2] Definitions and Explanations. United States Census. https://www.census.gov/housing/hvs/definitions.pdf 

\
Home Ownership as a voting issue:

[3] O’Donnell, K. (2022, August 2). Young people are pissed off: Housing crush sours millennial voters. Politico. https://www.politico.com/news/2022/08/02/housing-millennials-biden-economy-00047704  

\
Historical Intrest Rates:

[4] Freddie Mac, 30-Year Fixed Rate Mortgage Average in the United States [MORTGAGE30US], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/MORTGAGE30US, April 26, 2024.

\
Historical Median Home Price:

[5] https://dqydj.com/historical-home-prices/ 

\
Geojson for states:

https://rstudio.github.io/leaflet/json/us-states.geojson

\
Geojson for counties:

https://gist.github.com/sdwfrost/d1c73f91dd9d175998ed166eb216994a